\MakeUppercase{Layered Partitions of $k$-Planar Graphs} \thanks{This work was partly funded by NSERC and MRI.}

### 2019-06-17 k-planar.tex
Added proof for k-NN graphs

We conclude by pointing out two other graph classes, namely $(g,d)$-map graphs and $(g,\delta)$-string graphs that are $(g,k)$-planar for $k=O(d^2)$ and $k=\delta^2$, respectively.  This implies that Theorems~\ref{thm:k-planar}--\ref{thm:g-k-planar} and Corollaries~\ref{cor:first}--\ref{cor:last} apply to $(g,d)$-map graphs and to $(g,\delta)$-string graphs.

@@

>>>>

We conclude by pointing out three other graph classes, namely $(g,d)$-map graphs, $(g,\delta)$-string graphs, and $k$-nearest-neighbour graphs that are $(g,k)$-planar for $k=O(d^2)$ and $k=\delta^2$, respectively.  This implies that Theorems~\ref{thm:k-planar}--\ref{thm:g-k-planar} and Corollaries~\ref{cor:first}--\ref{cor:last} apply to $(g,d)$-map graphs, $(g,\delta)$-string graphs, and $k$-nearest-neighbour graphs.

@@

\section{$k$-Nearest-Neighbour Graphs}

In this section, we show that $k$-Nearest neighbour graphs of point in the plane are $O(k^2)$-planar.  For two points $x,y\in\R^2$, let $d_2(x,y)$ denote the Euclidean distance between $x$ and $y$. The $k$-nearest-neighbour graph of a point set $P\subset\R^2$ is the geometric graph $G$ with vertex set $V(G)=P$ and edge set
\[  
  E= \{vw : v,w\in V(G),\, |\{x\in P\setminus\{v\}: d_2(v,x)\le d_2(v,w)\}|\le k\} \enspace .
\] 
(The edges of $G$ are straight-line segments joining their endpoints.)

The following result, which is immediate from \citet[Corollary~4.2.6]{abrego.munroy.ea:on} states that $k$-nearest-neighbour graphs have bounded maximum degree:
\begin{lem}\lemlabel{k-nn-max-degree}
  The degree of any vertex in a $k$-nearest-neighbour graph is at most $6k$.
\end{lem}

We make use of the following well-known observation (see e.g., \citet[Lemma~2]{bose.morin.ea:routing}):
\begin{obs}\obslabel{convex}
  If $v_0,\ldots,v_3$ are the vertices of a convex quadrilateral in counterclockwise order then there exists at least one $i\in\{0,\ldots,3\}$ such that $\max\{d_2(v_i,v_{i-1}), d_2(v_i,v_{i+1})\} < d_2(v_{i-1},v_{i+1})$, where subscripts are taken modulo 4.
\end{obs}

\begin{lem}
  Every $k$-nearest-neighbour graph is $O(k^2)$ planar.
\end{lem}

\begin{proof}
  Let $G$ be a $k$-nearest-neighbour graph and consider any edge $vw\in E(G)$.
  Let $xy\in E(G)$ be an edge that crosses $vw$.  Note that $vxwy$ are the vertices of a convex quadrilateral in (without loss of generality) counterclockwise order. Then we say that 
  \begin{compactenum}
    \item $xy$ is of Type~$v$ if $\max\{d_2(v,x), d_2(v,y)\}< d_2(x,y)$;
    \item $xy$ is of Type~$w$ if $\max\{d_2(w,x), d_2(w,y)\}< d_2(x,y)$; or
    \item $xy$ is of Type~C otherwise.
  \end{compactenum}
  If $xy$ is of Type~C, then \obsref{convex} implies that (without loss of generality) $\max\{d_2(x,v),d_2(x,w)\} < d_2(v,w)$.  In this case, we call $x$ a Type~C vertex.  We claim that $V(G)$ contains at most $k-1$ Type~C vertices.  Indeed, more than $k-1$ Type~C vertices would contradict the fact that $vw\in E(G)$ since every Type~C vertex is closer to both $v$ and $w$ than $d_2(v,w)$.

  Next oberve that, if $xy$ is of Type~$v$, then at least one of $xv$ or $yv$ is in $E(G)$ in which case we call $x$ (respectively $y$) a Type~$v$ vertex.  By \lemref{k-nn-max-degree}, there are at most $6k$ Type~$v$ vertices.  Similarly, there are at most $6k$ Type~$w$ vertices.

  Thus, in all, there are at most $13k-1$ Type~$v$, Type~$w$, and Type~C vertices in total. By \lemref{k-nn-max-degree}, each of these vertices is incident on at most $6k$ edges that cross $vw$. Therefore, there are at most $78k^2-6k$ edges of $G$ that cross $vw$.  Since this is true for every edge $vw\in E(G)$, $G$ is $(78k^2-6k)$-planar.
\end{proof}
